An empirical assessment of strategies to model opponent effects in crash severity analysis

### 2020-01-23 Interactions-Crash-Severity-v1/Interactions-Crash-Severity-v1.tex
Completed Section: Setting for empirical assessment

% Pandoc header
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Road safety continues to be a concern world-wide. According to a recent
report from the World Health Organization (2019), road accidents are the
8th leading cause of death for all ages, and the number one cause of
death for children and young people between the ages of 5 to 29. Of all
leading causes of death, road accidents are the only cause of death
unrelated to disease, disorder, or infection. Road accidents impose a
heavy burden on individuals and society as a whole. Gobally, the rate of
road collision-related deaths per 100,000 population and 100,000
vehicles have both fallen, even as the number of vehicles has grown
(World Health Organization, 2019, Figs. 1 and 2). These gains, although
they are to be celebrated, cannot distract from the crushing economic
cost of premature death (e.g., Symons et al., 2019; Wijnen et al.,
2019), not to mention the long-term consequences for survivors, measured
in sometimes crippling emotional and physical pain (e.g., Merlin et al.,
2007; Devlin et al., 2019; Pelissier et al., n.d.).

@@

Table \ref{tab:model-summary} collects some key summary statistics of
the estimated models. Of interest is the goodness of fit of the models,
which in the case is measured with Akaike's Information Criterion (AIC).
This criterion is calculated using the estimated likelihood of the
model, penalized by the number of coefficients in the model, so that
other things being equal, it gives preference to more parsimonious
models. Since smaller values of AIC are better fits, it is possible to
see that, compared to the base model without opponent variables (Model
1), there are large and significant improvements in fit to be gained by
introducing opponent effects. However, the gains are not as large when
hierarchical specifications are used, even when the number of additional
coefficients that need to be estimated is not substantially larger
(recall that the penalty per parameter in AIC is 2).

The likelihood function of the model, and therefore the value of the
AIC, depend on the size of the sample, which is why AIC is not
comparable across models estimated with different sample sizes. For this
reason, the full sample models cannot be compared directly to the models
estimated with subsets of data. The models in the ensembles, however,
can be compared, and there we find that introducing opponent variables
leads to a better fit in most, but not all cases. One model for
pedestrians (when the opponent is a heavy vehicle), the two models for
bicyclists, and the three models for motorcyclists are better fits when
opponent variables are not used.

@@

\caption{\label{tab:model-summary}\label{tab:model-summary}Summary of model estimation results}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{lccc}
\toprule
Model & \makecell[l]{Number of\\observations} & \makecell[l]{Number of\\coefficients} & AIC\\
\midrule
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Full sample models}}\\
\hspace{1em}Model 1 & 164,511 & 104 & 195,215\\
\hspace{1em}Model 2 & 164,511 & 110 & 178,943\\
\rowcolor{gray!6}  \hspace{1em}Model 3 & 164,511 & 120 & 181,333\\
\hspace{1em}Model 4 & 164,511 & 113 & 179,018\\
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Model 1 Ensemble (sample subsets by user type vs opponent)}}\\
\hspace{1em}Light duty vehicle vs light duty vehicle & 114,841 & 96 & 145,390\\
\hspace{1em}Light duty vehicle vs light truck & 3,237 & 81 & 3,943\\
\rowcolor{gray!6}  \hspace{1em}Light duty vehicle vs heavy vehicle & 5,013 & 90 & 5,895\\
\hspace{1em}Light truck vs light duty vehicle & 3,121 & 81 & 3,885\\
\rowcolor{gray!6}  \hspace{1em}Light truck vs light truck & 809 & 69 & 1,170\\
\hspace{1em}Light truck vs heavy vehicle & 198 & 66 & 288\\
\rowcolor{gray!6}  \hspace{1em}Heavy vehicle vs light duty vehicle & 4,726 & 81 & 4,326\\
\hspace{1em}Heavy vehicle vs light truck & 180 & 66 & 225\\
\rowcolor{gray!6}  \hspace{1em}Heavy vehicle vs heavy vehicle & 779 & 76 & 1,147\\
\hspace{1em}Pedestrian vs light duty vehicle & 7,176 & 90 & 2,826\\
\rowcolor{gray!6}  \hspace{1em}Pedestrian vs light truck & 328 & 64 & 202\\
\hspace{1em}Pedestrian vs heavy vehicle & 376 & 66 & 409\\
\rowcolor{gray!6}  \hspace{1em}Bicyclist vs light duty vehicle & 3,521 & 82 & 654\\
\hspace{1em}Bicyclist vs light truck & 116 & 44 & 84\\
\rowcolor{gray!6}  \hspace{1em}Bicyclist vs heavy vehicle & NA & NA & \vphantom{1} NA\\
\hspace{1em}Motorcyclist vs light duty vehicle & 2,298 & 80 & 1,367\\
\rowcolor{gray!6}  \hspace{1em}Motorcyclist vs light truck & 127 & 58 & 153\\
\hspace{1em}Motorcyclist vs heavy vehicle & 62 & 45 & 88\\
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Model 2 Ensemble (sample subsets by user type vs opponent)}}\\
\hspace{1em}Light duty vehicle vs light duty vehicle & 114,841 & 99 & 143,903\\
\hspace{1em}Light duty vehicle vs light truck & 3,237 & 84 & 3,927\\
\rowcolor{gray!6}  \hspace{1em}Light duty vehicle vs heavy vehicle & 5,013 & 93 & 5,878\\
\hspace{1em}Light truck vs light duty vehicle & 3,121 & 84 & 3,877\\
\rowcolor{gray!6}  \hspace{1em}Light truck vs light truck & 809 & 72 & 1,156\\
\hspace{1em}Light truck vs heavy vehicle & 198 & 69 & 281\\
\rowcolor{gray!6}  \hspace{1em}Heavy vehicle vs light duty vehicle & 4,732 & 86 & 4,283\\
\hspace{1em}Heavy vehicle vs light truck & 179 & 67 & 205\\
\rowcolor{gray!6}  \hspace{1em}Heavy vehicle vs heavy vehicle & 779 & 79 & 1,136\\
\hspace{1em}Pedestrian vs light duty vehicle & 7,176 & 93 & 2,821\\
\rowcolor{gray!6}  \hspace{1em}Pedestrian vs light truck & 328 & 67 & 200\\
\hspace{1em}Pedestrian vs heavy vehicle & 376 & 69 & 410\\
\rowcolor{gray!6}  \hspace{1em}Bicyclist vs light duty vehicle & 3,521 & 85 & 659\\
\hspace{1em}Bicyclist vs light truck & 145 & 59 & 114\\
\rowcolor{gray!6}  Bicyclist vs heavy vehicle & NA & NA & \vphantom{1} NA\\
\hspace{1em}Motorcyclist vs light duty vehicle & 2,298 & 83 & 1,373\\
\rowcolor{gray!6}  \hspace{1em}Motorcyclist vs light truck & 127 & 61 & 153\\
\hspace{1em}Motorcyclist vs heavy vehicle & 63 & 45 & 90\\
\bottomrule
\multicolumn{4}{l}{\textit{Note: }}\\
\multicolumn{4}{l}{There are zero cases of Bicyclist vs heavy vehicle in the sample}\\
\end{tabular}
\end{table}

@@

\hypertarget{outcome-shares-based-on-estimated-probabilities}{%
\subsection{Outcome shares based on estimated
probabilities}\label{outcome-shares-based-on-estimated-probabilities}}

@@

\hypertarget{predicted-outcomes}{%
\subsection{Predicted outcomes}\label{predicted-outcomes}}

@@

\hypertarget{sec:insights}{%
\section{Best model: insights}\label{sec:insights}}

Create a table with the results of Model 2.

>>>>

% Pandoc header
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Road safety continues to be a world-wide concern. According to a recent
report from the World Health Organization (2019), road accidents are the
8th leading cause of death for all ages, and the number one cause of
death for children and young people between the ages of 5 to 29. Of all
leading causes of death, road accidents are the only cause of death
unrelated to disease, disorder, or infection. Road accidents impose a
heavy burden on individuals and society as a whole. Gobally, the rate of
road collision-related deaths per 100,000 population and 100,000
vehicles have both fallen, even as the number of vehicles has grown
(World Health Organization, 2019, Figs. 1 and 2). These gains, although
they are to be celebrated, cannot distract from the crushing economic
cost of premature death (e.g., Symons et al., 2019; Wijnen et al.,
2019), not to mention the long-term consequences for survivors, measured
in sometimes crippling emotional and physical pain (e.g., Merlin et al.,
2007; Devlin et al., 2019; Pelissier et al., n.d.).

@@

Tables \ref{tab:model-full-sample-summary} and
\ref{tab:model-subsample-summary} collect some key summary statistics of
the estimated models. Of interest is the goodness of fit of the models,
which in the case is measured with Akaike's Information Criterion (AIC).
This criterion is calculated using the estimated likelihood of the
model, penalized by the number of coefficients in the model, so that,
other things being equal, it gives preference to more parsimonious
models. The objective is to minimize the AIC, and therefore smaller
values of this criterion represent better model fits. Turning our
attention to the models estimated using the full sample (Table
\ref{tab:model-full-sample-summary}), it is possible to see that,
compared to the base (single level) model without opponent variables
(Model 1), there are large and significant improvements in fit to be
gained by introducing opponent effects. However, the gains are not as
large when hierarchical specifications are used, even when the number of
additional coefficients that need to be estimated is not substantially
larger (recall that the penalty per parameter in AIC is 2). The best
model according to this measure of goodness of fit is Model 2
(single-level with opponent effects), followed by Model 4 (hierarchical
opponent variables), Model 3 (hierarchical traffic unit variables with
opponent effects), and finally Model 1 (single level without opponent
effects).

It is important to note that the likelihood function of the model, and
therefore the value of the AIC, both depend on the size of the sample,
which is why AIC is not comparable across models estimated with
different sample sizes. For this reason, the full sample models cannot
be compared directly to the models estimated with subsets of data. The
models in the ensembles, however, can be compared to each other (Table
\ref{tab:model-subsample-summary}). As seen in the table, introducing
opponent variables leads to a better fit in the case of most, but not
all models. The simplest model (single level without opponent effects)
is clearly the best fitting candidate in the case of bicycle vs light
truck collisions, bicycle vs heavy vehicle collisions, motorcyclist vs
light duty vehicle collisions, and motorcyclist vs heavy vehicle
collisions. Model 1 is a statistical toss for best performance with
other model(s) in the case of pedestrian vs heavy vehicle collisions.
The relative likelihood of Model 1 compared to Models 2 and 3 in this
case is 0.56, which means that either of these two competing models are
0.56 times as probable as Model 1 to minimize the information loss.

Model 2 is the best fit in the case of light truck vs heavy vehicle
collisions. This model is also tied for best fit with Model 2 in the
case of pedestrian vs light duty vehicle and pedestrian vs light truck
collisions, and is a statistical toss with Model 4 in the case of heavy
vehicle vs heavy vehicle collisions (relative likelihood is 0.592).
Model 3 is the best fit in the case of light duty vehicle vs light duty
vehicle collisions and heavy vehicle vs light duty vehicle. Model 4 is
the best fit in the case of light duty vehicle vs light truck
collisions, light duty vehicle vs heavy vehicle collisions, light truck
vs light duty vehicle collisions, heavy vehicle vs light truck
collisions, and motorcycle vs light truck collisions. This model is a
statistical toss with Model 2 in the case of light truck vs light truck
collisions, with a relative likelihood of 0.791.

@@

\caption{\label{tab:model-full-sample-summary}\label{tab:model-full-sample-summary}Summary of model estimation results: Full sample models}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{lccc}
\toprule
Model & \makecell[l]{Number of\\observations} & \makecell[l]{Number of\\coefficients} & AIC\\
\midrule
\rowcolor{gray!6}  Model 1 & 164,511 & 102 & 195,215\\
Model 2 & 164,511 & 108 & 178,943\\
\rowcolor{gray!6}  Model 3 & 164,511 & 118 & 181,333\\
Model 4 & 164,511 & 111 & 179,018\\
\bottomrule
\end{tabular}
\end{table}

\caption{\label{tab:model-subsample-summary}\label{tab:model-summary}Summary of model estimation results: Subsample Models}
\centering
\resizebox{\linewidth}{!}{
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{lccccccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2} & \multicolumn{2}{c}{Model 3} & \multicolumn{2}{c}{Model 4} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10}
Model & \makecell[l]{Number of\\observations} & \makecell[l]{Number of\\coefficients} & AIC & \makecell[l]{Number of\\coefficients} & AIC & \makecell[l]{Number of\\coefficients} & AIC & \makecell[l]{Number of\\coefficients} & AIC\\
\midrule
\rowcolor{gray!6}  Light duty vehicle vs light duty vehicle & 114,841 & 94 & 145,390 & 97 & 143,903 & 100 & 143,896 & 100 & 144,004\\
Light duty vehicle vs light truck & 3,237 & 79 & 3,943 & 82 & 3,927 & 85 & 3,937 & 85 & 3,922\\
\rowcolor{gray!6}  Light duty vehicle vs heavy vehicle & 5,013 & 88 & 5,895 & 91 & 5,878 & 94 & 5,888 & 94 & 5,864\\
Light truck vs light duty vehicle & 3,121 & 79 & 3,885 & 82 & 3,877 & 85 & 3,881 & 85 & 3,875\\
\rowcolor{gray!6}  Light truck vs light truck & 809 & 67 & 1,170 & 70 & 1,156 & 73 & 1,162 & 73 & 1,155\\
\addlinespace
Light truck vs heavy vehicle & 198 & 64 & 288 & 67 & 281 & 70 & 287 & 70 & 286\\
\rowcolor{gray!6}  Heavy vehicle vs light duty vehicle & 4,726 & 79 & 4,326 & 84 & 4,283 & 86 & 4,268 & 87 & 4,287\\
Heavy vehicle vs light truck & 180 & 64 & 225 & 65 & 205 & 67 & 207 & 66 & 187\\
\rowcolor{gray!6}  Heavy vehicle vs heavy vehicle & 779 & 74 & 1,147 & 77 & 1,136 & 80 & 1,141 & 80 & 1,137\\
Pedestrian vs light duty vehicle & 7,176 & 88 & 2,826 & 91 & 2,821 & 91 & 2,821 & 93 & 2,827\\
\addlinespace
\rowcolor{gray!6}  Pedestrian vs light truck & 328 & 62 & 202 & 65 & 200 & 65 & 200 & 68 & 206\\
Pedestrian vs heavy vehicle & 376 & 64 & 409 & 67 & 410 & 67 & 410 & 70 & 417\\
\rowcolor{gray!6}  Bicyclist vs light duty vehicle & 3,521 & 80 & 654 & 83 & 659 & 83 & 659 & 85 & 657\\
Bicyclist vs light truck & 116 & 42 & 84 & 57 & 114 & 57 & 114 & 54 & 108\\
\rowcolor{gray!6}  Bicyclist vs heavy vehicle & - & - & - & - & - & - & - & - & -\\
\addlinespace
Motorcyclist vs light duty vehicle & 2,298 & 78 & 1,367 & 81 & 1,373 & 81 & 1,373 & 84 & 1,373\\
\rowcolor{gray!6}  Motorcyclist vs light truck & 127 & 56 & 153 & 59 & 153 & 59 & 153 & 47 & 94\\
Motorcyclist vs heavy vehicle & 62 & 43 & 88 & 45 & 90 & 46 & 92 & 51 & 102\\
\bottomrule
\multicolumn{10}{l}{\textit{Note: }}\\
\multicolumn{10}{l}{There are zero cases of Bicyclist vs heavy vehicle in the sample}\\
\end{tabular}}
\end{table}
\end{landscape}

@@

\hypertarget{outcome-shares-based-on-predicted-probabilities}{%
\subsection{Outcome shares based on predicted
probabilities}\label{outcome-shares-based-on-predicted-probabilities}}

@@

\hypertarget{outcome-frequency-based-on-predicted-classes}{%
\subsection{Outcome frequency based on predicted
classes}\label{outcome-frequency-based-on-predicted-classes}}

@@
